### 2018-03-12 paper/3-compute.tex
Minor revision (typos, TODOs)

\noindent
\hs{Compute} represents an abstract set of \emph{build tasks}, for example, build
rules in software build systems or formulas in spreadsheets. It is completely
isolated from the world of compilers, file systems, calculation chains, caches,
and all other complexities of real build systems. We explain the compute type
in~\S\ref{sec-general-compute}, discussing several noteworthy special cases:
functorial, applicative and monadic types of compute.

@@

We can also execute a compute on a given store to obtain the resulting value.
This uses a very similar transformation based on the \hs{Identity} monad
defined in \cmd{Data.Functor.Identity}:

>>>>

\noindent
\hs{Compute} represents an abstract set of \emph{build tasks}, for example, build
rules in software build systems or formulas in spreadsheets. It is completely
isolated from the world of compilers, file systems, calculation chains, caches,
and all other complexities of real build systems. We explain the compute type
in~\S\ref{sec-compute-applicative}, discussing several noteworthy special cases:
functorial, applicative and monadic types of compute.

@@

We can also execute a compute on a given store to obtain the resulting value.
This uses a similar transformation based on the \hs{Identity} monad
defined in \cmd{Data.Functor.Identity}:

### 2018-03-12 paper/5-implementations.tex
Minor revision (typos, TODOs)

\todo{AM}{AM: What about using shorter \hs{get} and \hs{put}? I can introduce it right in the common vocabulary, mentioning the usual get-put laws. I would also remove `get' prefixes from many functions, e.g. \hs{dependencies compute} reads fine. I suggest not to rush with this TODO -- let me first finish the Compute section, where many of these helper functions will be defined. NM: I think this is a bad idea - conflicts with a state Monad which is what we're using underneath so will confuse some readers.}

@@

We'll come back to \hs{withChanged} afterwards, but ignoring that, reading the rest of the implementation you can see the underying essence of \Make. We use the helper function \hs{topological} which gets called with each file in a dependency-respecting order given the file (\hs{k}), the dependencies (\hs{dk}) and an action to generate the result (\hs{act}). With those three things in hand, the implementation first gets the timestamp of the key being built, then the timestamps of the dependencies (which must have been rebuilt if necessary, thanks to the guarantees on \hs{topological}), and considers it dirty if any dependency was built after \hs{k}. If it is dirty, it rebuilds and stores the result.

@@

\begin{minted}{haskell}
topological
    :: (m ~ M i k v, Default i)
    => (k -> [k] -> m v -> m ())
    -> Build Applicative i k v
topological step compute k = runM $ do
    let depends = getDependencies compute
    forM_ (topSort depends $ transitiveClosure depends k) $ \k ->
        case compute getStore k of
            Nothing -> return ()
            Just act -> step k (depends k) act
\end{minted}

It uses a \hs{getDependencies} on the computation function, takes the transitive-closure and topological-sort to provide an order in which to build the keys. For each key, if it is an input it does nothing, otherwise it calls the supplied \hs{step} function. Note that \hs{getDependencies} is only defined for \hs{Applicative} computes, which is what restricts \Make to Applicative build systems. Moreover, anyone following the same topological-sort approach will also have the \hs{Applicative} restriction.

The greatest impedence mismatch between our framework and this implementation is the notion of \hs{Time} in \hs{getStoreTime}, which is missing from \hs{Build}. However, as we discussed in \S\ref{sec-dirty-bit}, \Make is really using modification-time to encode a dirty bit with specific properties. We model the dirty bit and time with the following API on \hs{M}:

@@

We use the helper function \hs{reordering} which provides the key to build (\hs{k}), the dependencies of \hs{k} (\hs{dk}) and an action to build \hs{k} (\hs{act}). The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Compute} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway. If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possiblities:

@@

The astute reader may notice that \hs{reordering} supplies extra information that is ignored with \hs{_} patterns. While not required for \Excel, we have implemented build systems using \hs{reordering} which use verifying and constructive traces, effectively turning \Excel into a cloud build system and ensuring \hs{reordering} is not overly fitted to \Excel alone (additional implementations available in the suplementary material). The definition of \hs{reordering} is:

@@

\begin{enumerate}
\item \hs{k} - the key that is being built.
\item \hs{Maybe [k]} - the list of dependencies, or \hs{Nothing} if the step makes use of monadic dependencies. Note that above if a step is truely monadic it is always rebuilt. In the case of spreadsheets such steps are very rare, so a very conservative approximation can be used.\footnote{In truth, many spreadsheets take the alternative approximation of always assuming monadic steps \textit{did not} change - trading correctness for performance.}
\item \hs{k -> m (Maybe v)} - a function that looks up a key - it returns \hs{Nothing} if that key has not yet been built (and thus there is a dependency violation).
\item \hs{m (Either k ([k], v))} - an action to produce either the \hs{k} which is required but not yet computed (the reason you fail your dependency order check), or a list of the keys used and value produced.
\item It returns \hs{m (Maybe k)}, with a \hs{Nothing} to indicate the the build succeeded, or \hs{Just k} to indicate that the given key \hs{k} has not been built, and thus this key should be moved after \hs{k}.
\end{enumerate}

@@

This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to fetch the result of a dependecy (\hs{fetch}), and a way to build \hs{k} (\hs{act}). The code first looks up the traces that match the current situation, creating a list of possible results as \hs{poss}, using the helper \hs{traceMatch}. We check if the current value for the key is one of the possible results, and it not, build the key and store it in the trace. We can implement \hs{traceMatch}, which is used in all build systems working with \hs{Trace}, as:

@@

The required \hs{step} function gets given the \hs{k} to build, the dependencies as \hs{Maybe [k]} (where they are \hs{Nothing} if the rule is truely monadic), a function to demand the result of a dependent key, and a function to compute the current value and return the actual dependencies. To ensure each file is only visited once in a single execution a piece of temporary state is used with \hs{getTemp} and \hs{modifyTemp} to record which files were visited - this state is not persisted to the \hs{i} info parameter and is reset on each fresh execution of \hs{M}.

@@

Now we have seen the three dependency schemes, we can directly reuse \hs{topological} to define Bazel. Furthermore, as Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type. We define \Bazel as:

@@

We execute keys in a topological order. For each key we first get a list of the traces that match. If there are none we rebuild and record the trace, much like \Shake, but also recording the value of the result. If there are some matching traces, and the current value of \hs{k} is not compatible, we grab the first possiblity and grab its value from \hs{contents}.

@@

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}. Therefore, such systems can store \hs{Map k (Trace k v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
\item Any \hs{Applicative} system storing constructive traces, e.g. \Bazel, can index directly from the key and results to the output result - i.e. \hs{Map (k, [Hash v]) (Hash v)}. Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash. In this formulation we have removed the possibilty for a single key/dependency state to map to multiple different hashes, e.g. on a non-determinstic build - something \Bazel already prohibits which is discussed more in \S\ref{sec-non-determinism}.
\item Finally, a \hs{Monad} system with constructive traces can be efficiently encoded as \hs{Map k (Choice k v)}, assuming a definition of \hs{Choice} as:
\begin{minted}{haskell}
data Choice k v = Choice k (Map (Hash v) Choice)
                | Result (Hash v)
\end{minted}
Here the \hs{Choice} effectively encodes a tree, asking successive questions about keys, and taking different branches based on the resulting value, until it reaches a final result. Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
\end{enumerate}

>>>>

@@

We'll come back to \hs{withChanged} afterwards, but ignoring that, reading the rest of the implementation you can see the underlying essence of \Make. We use the helper function \hs{topological} which gets called with each file in a dependency-respecting order given the file (\hs{k}), the dependencies (\hs{dk}) and an action to generate the result (\hs{act}). With those three things in hand, the implementation first gets the timestamp of the key being built, then the timestamps of the dependencies (which must have been rebuilt if necessary, thanks to the guarantees on \hs{topological}), and considers it dirty if any dependency was built after \hs{k}. If it is dirty, it rebuilds and stores the result.

@@

\begin{minted}{haskell}
topological
    :: (m ~ M i k v, Default i)
    => (k -> [k] -> m v -> m ())
    -> Build Applicative i k v
topological step compute k = runM $ do
    let depends = dependencies compute
    forM_ (topSort depends $ transitiveClosure depends k) $ \k ->
        case compute getStore k of
            Nothing -> return ()
            Just act -> step k (depends k) act
\end{minted}

It uses a \hs{dependencies} on the computation function, takes the transitive-closure and topological-sort to provide an order in which to build the keys. For each key, if it is an input it does nothing, otherwise it calls the supplied \hs{step} function. Note that \hs{dependencies} is only defined for \hs{Applicative} computes, which is what restricts \Make to Applicative build systems. Moreover, anyone following the same topological-sort approach will also have the \hs{Applicative} restriction.

The greatest impedance mismatch between our framework and this implementation is the notion of \hs{Time} in \hs{getStoreTime}, which is missing from \hs{Build}. However, as we discussed in \S\ref{sec-dirty-bit}, \Make is really using modification-time to encode a dirty bit with specific properties. We model the dirty bit and time with the following API on \hs{M}:

@@

We use the helper function \hs{reordering} which provides the key to build (\hs{k}), the dependencies of \hs{k} (\hs{dk}) and an action to build \hs{k} (\hs{act}). The dirty check is the same as \hs{makeDirtyBit}, with the added complication that because the \hs{Compute} is monadic it may not be possible to statically get a list of dependencies - if \hs{dk} is \hs{Nothing} then the action uses the monadic power and we conservatively rebuild anyway. If the check determines that the key is dirty, and needs rebuilding, we run \hs{act}, giving two possibilities:

@@

The astute reader may notice that \hs{reordering} supplies extra information that is ignored with \hs{_} patterns. While not required for \Excel, we have implemented build systems using \hs{reordering} which use verifying and constructive traces, effectively turning \Excel into a cloud build system and ensuring \hs{reordering} is not overly fitted to \Excel alone (additional implementations available in the supplementary material). The definition of \hs{reordering} is:

@@

\begin{enumerate}
\item \hs{k} - the key that is being built.
\item \hs{Maybe [k]} - the list of dependencies, or \hs{Nothing} if the step makes use of monadic dependencies. Note that above if a step is truly monadic it is always rebuilt. In the case of spreadsheets such steps are very rare, so a very conservative approximation can be used.\footnote{In truth, many spreadsheets take the alternative approximation of always assuming monadic steps \textit{did not} change - trading correctness for performance.}
\item \hs{k -> m (Maybe v)} - a function that looks up a key - it returns \hs{Nothing} if that key has not yet been built (and thus there is a dependency violation).
\item \hs{m (Either k ([k], v))} - an action to produce either the \hs{k} which is required but not yet computed (the reason you fail your dependency order check), or a list of the keys used and value produced.
\item It returns \hs{m (Maybe k)}, with a \hs{Nothing} to indicate the the build succeeded, or \hs{Just k} to indicate that the given key \hs{k} has not been built, and thus this key should be moved after \hs{k}.
\end{enumerate}

@@

This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to fetch the result of a dependency (\hs{fetch}), and a way to build \hs{k} (\hs{act}). The code first looks up the traces that match the current situation, creating a list of possible results as \hs{poss}, using the helper \hs{traceMatch}. We check if the current value for the key is one of the possible results, and it not, build the key and store it in the trace. We can implement \hs{traceMatch}, which is used in all build systems working with \hs{Trace}, as:

@@

The required \hs{step} function gets given the \hs{k} to build, the dependencies as \hs{Maybe [k]} (where they are \hs{Nothing} if the rule is truly monadic), a function to demand the result of a dependent key, and a function to compute the current value and return the actual dependencies. To ensure each file is only visited once in a single execution a piece of temporary state is used with \hs{getTemp} and \hs{modifyTemp} to record which files were visited - this state is not persisted to the \hs{i} info parameter and is reset on each fresh execution of \hs{M}.

@@

Now we have seen the three dependency schemes, we can directly reuse \hs{topological} to define \Bazel. Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type. We define \Bazel as:

@@

We execute keys in a topological order. For each key we first get a list of the traces that match. If there are none we rebuild and record the trace, much like \Shake, but also recording the value of the result. If there are some matching traces, and the current value of \hs{k} is not compatible, we grab the first possibility and grab its value from \hs{contents}.

@@

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see significant benefit from storing more than one \hs{Trace} per key\footnote{There is a small chance of a benefit if the dependencies change but the result does not, and then the dependencies change back to what they were before.}. Therefore, such systems can store \hs{Map k (Trace k v)}, where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
\item Any system using \hs{Applicative} dependencies can omit the dependency keys from the \hs{Trace} as they can be recovered from the \hs{key} field.
\item Any \hs{Applicative} system storing constructive traces, e.g. \Bazel, can index directly from the key and results to the output result - i.e. \hs{Map (k, [Hash v]) (Hash v)}. Importantly, assuming the traces are stored on a central server, the client can compute the key and the hashes of its dependencies, then make a single call to the server to retrieve the result hash. In this formulation we have removed the possibility for a single key/dependency state to map to multiple different hashes, e.g. on a non-deterministic build - something \Bazel already prohibits which is discussed more in \S\ref{sec-non-determinism}.
\item Finally, a \hs{Monad} system with constructive traces can be efficiently encoded as \hs{Map k (Choice k v)}, assuming a definition of \hs{Choice} as:
\begin{minted}{haskell}
data Choice k v = Choice k (Map (Hash v) Choice)
                | Result (Hash v)
\end{minted}
Here the \hs{Choice} effectively encodes a tree, asking successive questions about keys, and taking different branches based on the resulting value, until it reaches a final result. Implementing this structure over client-server communication requires either a chatty interface with lots of round-trips per \hs{Choice} step, or sending over a part of the tree that is not subsequently explored.
\end{enumerate}

### 2018-03-12 paper/6-engineering.tex
Minor revision (typos, TODOs)

Interestingly, there is signficant interplay between all three sources of impurity. Systems like \Bazel use various sandboxing techniques to guard against missing dependencies, but none are likely to capture all dependencies right down to CPU model and microcode version. Rules that do have untracked state can be marked as volatile, a technique \Excel takes with the \textsf{INDIRECT} function, removing the untracked-state at the cost of more recomputation.

@@

The build systems that have precise tracking of self-computation are all onces which use a non-Embedded Domain Specific Language to describe computations. Those which make use of a full programming language, e.g. \Shake, are much more difficult to implement equality on rules as they are arbitrary functions making use of opqaue abstractions. For such build systems the incredibly pessimistic assuming of saying that any change to the build system potentially changes any rule can often be used -- the classic example being a \texttt{Makefile} depending on itself.

@@

For examples like \LaTeX we consider the proper encoding to not be with circular rules, but with a series of iterative steps, as described in \S\ref{neil-iterative-on-stack-overflow}. It is important that the number of executions is bounded, otherwise the build system may not terminate (a legitimate concern with \LaTeX which can be put into a situation where it is bistable or divering over multiple executions).

@@

Our build system abstraction assumes a \hs{k -> v} store, along with a build system that works directly on \hs{k} and \hs{v} values. However, certain build systems provide greater flexiblity, e.g. \Shake permits polymorphic keys and values, allowing types that are only stored in the Shake information, and never persisted to the Store.

>>>>

Interestingly, there is significant interplay between all three sources of impurity. Systems like \Bazel use various sandboxing techniques to guard against missing dependencies, but none are likely to capture all dependencies right down to CPU model and microcode version. Rules that do have untracked state can be marked as volatile, a technique \Excel takes with the \textsf{INDIRECT} function, removing the untracked-state at the cost of more recomputation.

@@

The build systems that have precise tracking of self-computation are all ones which use a non-Embedded Domain Specific Language to describe computations. Those which make use of a full programming language, e.g. \Shake, are much more difficult to implement equality on rules as they are arbitrary functions making use of opaque abstractions. For such build systems the incredibly pessimistic assuming of saying that any change to the build system potentially changes any rule can often be used -- the classic example being a \texttt{Makefile} depending on itself.

@@

For examples like \LaTeX we consider the proper encoding to not be with circular rules, but with a series of iterative steps, as described in \S\ref{neil-iterative-on-stack-overflow}. It is important that the number of executions is bounded, otherwise the build system may not terminate (a legitimate concern with \LaTeX which can be put into a situation where it is bistable or diverging over multiple executions).

@@

Our build system abstraction assumes a \hs{k -> v} store, along with a build system that works directly on \hs{k} and \hs{v} values. However, certain build systems provide greater flexibility, e.g. \Shake permits polymorphic keys and values, allowing types that are only stored in the Shake information, and never persisted to the Store.

### 2018-03-12 paper/7-related.tex
Minor revision (typos, TODOs)

\begin{itemize}
\item \Ninja \cite{ninja} combines the dependency strategy of \Make with the validating traces of \Shake - our associated implementation provides such a combination. Ninja is also capable of modelling rules that produce multiple results, a limited form of polymorphism \S\ref{sec-polymorphism}.
\item \Tup \cite{tup} functions much like \Make, but with a refined dirty-bit implementation that watches the file system for changes and can thus avoid rechecking the entire graph, and with automatic deleting of stale results.
\item \Redo \cite{redo} almost exactly matches \Shake at the level of detail given here, differing only on aspects like polymorphic dependencies \S\ref{sec-polymorphism}.
\item \Buck \cite{buck} and \CloudBuild \cite{cloudbuild} are both fundamentally similar to \Bazel, differing around areas such as non-determinism \S\ref{sec-sec-non-determinism}.
\item \Nix \cite{nix} has coarse-grained monadic dependencies, with precise hashing of dependencies and downloading of precomputed build products, so may be, at some level of abstraction, the same as the hypothetical Cloud Shake from \S\ref{sec-implementation-cloud-shake}. However, \Nix is not intended as a build system, and the coarse grained nature (packages, not individual files) makes it targetted to a different purpose. \todo{NM}{Read comment below when not on an airplane.}
% John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
% \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.
\item \Pluto \cite{pluto} has power to express cyclic dependenies, but that is implemented \todo{NM}{Read the paper so I can say something precise}
\end{itemize}

@@

While not typically considered build systems, self-adjusting computation is a well studied area, and in particular the constrast between different formulations has been well studied. \todo{AM}{Say something.}

@@

Looks tanilisingly close to the profunctor optics definition in \cite{gibbons-profunctor-optics}:

@@

Assuming we instantiate \hs{p} to something that provides \hs{k -> f v} -- which many of the actual instances in that paper do. The properties of such optics are well studied, and the functions like \hs{getDependencies} are very much based on observations from that field of work. Alas, we have been unable to remove the \hs{Maybe} used to encode whether a file is an input, without complicating other aspects of our definition. Furthermore, the \hs{Build} operation lacks any further such symmetry.

>>>>

\begin{itemize}
\item \Ninja \cite{ninja} combines the dependency strategy of \Make with the validating traces of \Shake - our associated implementation provides such a combination. Ninja is also capable of modelling rules that produce multiple results, a limited form of polymorphism \S\ref{sec-polymorphism}.
\item \Tup \cite{tup} functions much like \Make, but with a refined dirty-bit implementation that watches the file system for changes and can thus avoid rechecking the entire graph, and with automatic deleting of stale results.
\item \Redo \cite{redo} almost exactly matches \Shake at the level of detail given here, differing only on aspects like polymorphic dependencies \S\ref{sec-polymorphism}.
\item \Buck \cite{buck} and \CloudBuild \cite{cloudbuild} are both fundamentally similar to \Bazel, differing around areas such as non-determinism \S\ref{sec-non-determinism}.
\item \Nix \cite{nix} has coarse-grained monadic dependencies, with precise hashing of dependencies and downloading of precomputed build products, so may be, at some level of abstraction, the same as the hypothetical Cloud Shake from \S\ref{sec-implementation-cloud-shake}. However, \Nix is not intended as a build system, and the coarse grained nature (packages, not individual files) makes it targeted to a different purpose. \todo{NM}{Read comment below when not on an airplane.}
% John Ericson suggested in a blog comment that nix may be somewhat monadic, see:
% \url{https://blogs.ncl.ac.uk/andreymokhov/cloud-and-dynamic-builds/\#comment-1849}.
\item \Pluto \cite{pluto} has power to express cyclic dependencies, but that is implemented \todo{NM}{Read the paper so I can say something precise}
\end{itemize}

@@

While not typically considered build systems, self-adjusting computation is a well studied area, and in particular the contrast between different formulations has been well studied. \todo{AM}{Say something.}

@@

Looks tantalisingly close to the profunctor optics definition in \cite{gibbons-profunctor-optics}:

@@

Assuming we instantiate \hs{p} to something that provides \hs{k -> f v} -- which many of the actual instances in that paper do. The properties of such optics are well studied, and the functions like \hs{dependencies} are very much based on observations from that field of work. Alas, we have been unable to remove the \hs{Maybe} used to encode whether a file is an input, without complicating other aspects of our definition. Furthermore, the \hs{Build} operation lacks any further such symmetry.
